---
title: "IDS investigation worksheet"
author: "by Just Commit: Liv, Max, Sarah, Helen, Andrew, & Adam"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_csv('Data/details_combined.csv')
fatalities = read_csv('Data/fatalities_combined.csv')
```

```{r Details Cleaning}
# Merge and convert formats of begin-date variables
details <- details %>%
  # Combine date elements
  unite("BEGIN_DATE", BEGIN_YEARMONTH, BEGIN_DAY, sep = "") %>% 
  mutate(
    # Convert date to a character for union with time
    BEGIN_DATE = as.character(BEGIN_DATE),
    # Add leading zero if the time is 3 digits
    BEGIN_TIME = case_when(
      BEGIN_TIME < 1000 ~ sprintf("%04d", BEGIN_TIME),
      TRUE ~ as.character(BEGIN_TIME)
    )
  ) %>%
  # Unite date and time into one string
  unite("BEGIN_DT", BEGIN_DATE, BEGIN_TIME, sep = " ") %>%
  mutate(
    # Convert to a datetime
    BEGIN_DT = ymd_hm(BEGIN_DT)
  )

# Merge and convert formats of end-date variables
details <- details %>%
  # Combine date elements
  unite("END_DATE", END_YEARMONTH, END_DAY, sep = "") %>% 
  mutate(
    # Convert date to a character for union with time
    END_DATE = as.character(END_DATE),
    # Add leading zero if the time is 3 digits
    END_TIME = case_when(
      END_TIME < 1000 ~ sprintf("%04d", END_TIME),
      TRUE ~ as.character(END_TIME)
    )
  ) %>%
  # Unite date and time into one string
  unite("END_DT", END_DATE, END_TIME, sep = " ") %>%
  mutate(
    # Convert to a datetime
    END_DT = ymd_hm(END_DT)
  )

# Remove unnecessary columns
details <- details %>%
  select(-c(YEAR, MONTH_NAME, BEGIN_DATE_TIME, END_DATE_TIME))

# Rename incorrect column names
details <- details %>%
  rename(REGION = STATE,
         REGION_FIPS = STATE_FIPS)

# Create duration variable 
details <- details %>% 
  mutate(
    duration = END_DT - BEGIN_DT
  ) %>% 
    mutate(
      duration = as.numeric(duration) / 60
    )

# Change cost suffixes
details <- details %>% 
  mutate(
    DAMAGE_PROPERTY = as.numeric(gsub("K", "e+03", gsub("M", "e+06", DAMAGE_PROPERTY))),
    DAMAGE_CROPS = as.numeric(gsub("K", "e+03", gsub("M", "e+06", DAMAGE_CROPS)))
  )

details <- details %>%
  mutate(
    DAMAGE_TOTAL = DAMAGE_PROPERTY + DAMAGE_CROPS
  )

details <- details %>% 
  filter(
    !year(BEGIN_DT) == '2024'
  )

details <- details %>% 
  clean_names()
```

```{r Fatalities Cleaning}
fatalities <- fatalities %>% 
  select(-FAT_TIME, -FATALITY_DATE, -EVENT_YEARMONTH) %>% 
  mutate(YMD = ymd(paste(FAT_YEARMONTH, FAT_DAY))) %>% 
  select(-FAT_DAY, -FAT_YEARMONTH)

fatalities <- fatalities %>% 
  clean_names()
```

```{r percentage damage plot, eval=FALSE}
cost_data <- details %>% 
  select(
    event_type, damage_property, damage_crops
  ) %>% 
    group_by(
      event_type
    ) %>% 
      summarise(
        Property = sum(damage_property, na.rm=TRUE),
        Crops = sum(damage_crops, na.rm=TRUE)
      ) %>%
        top_n(
          20, 
          Property + Crops
        ) %>% 
          pivot_longer(cols=c('Property', 'Crops'),
                      names_to='type',
                      values_to='damage'
          )

damage_plot <- cost_data %>% 
  ggplot(mapping=aes(
    x=damage/sum(damage)*100,
    y=reorder(event_type, damage),
    fill=type
  )) +
  geom_col() +
  labs(title = 'Damage by Storm Event Type between 1996 & 2023', 
       subtitle = 'Total Damage (Property + Crops) was $227bn',
       y = 'Type of Event',
       x = 'Pecentage of Total Damage',
       fill='Damage Type',
       caption='Only the 20 most damaging event types are shown.'
  ) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = -0.85, face= "italic"))

#ggsave('Damage by Event Type.jpg', plot=damage_plot)

```

```{r Simple Damage Statistics}
eval_percentage_criteria <- function(col, type) {
  col <- sym(col)
  
  if (type == 'event') {
    total <- details %>% 
      nrow()
  
    criteria_total <- details %>% 
      filter(
        !!col > 0
      ) %>%
        nrow()
    
    print('Percentage fulfilling criteria by event: ')
    print(criteria_total / total * 100)
  }
  
  else if (type == 'episode') {
    total <- details %>% 
      group_by(
        episode_id
      ) %>% 
        summarise(
          total = sum(!!col)
        ) %>% 
          nrow()
    
    criteria_total <- details %>% 
      group_by(
        episode_id
      ) %>% 
        summarise(
          total = sum(!!col)
        ) %>% 
          filter(
            total > 0
          ) %>%
            nrow()
    
    print('Percentage fulfilling criteria by episode: ')
    print(criteria_total / total * 100)
  }
    
  else {
    stop("Type must be one of 'event' or 'episode'")
  }
}
```

2.61% of storm episodes led to fatalities, while only 0.74% of events led to fatalities.
3.57% of storm episodes led to injuries, while only 1.02% of events led to injuries.
15.10% of storm episodes caused damage, while 15.39% of events causes damage.

```{r Who is most affected by weather}

fatalities %>%
  ggplot(mapping = aes(x = fatality_age)) +
    geom_histogram(binwidth = 4)

```

Going to add more details about what caused fatalities in different age groups and more - Andrew