---
title: "IDS investigation worksheet"
author: "by Just Commit: Liv, Max, Sarah, Helen, Andrew, & Adam"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_rds('data/clean_data/details_clean.rds')
fatalities = read_rds('data/clean_data/fatalities_clean.rds')
```

2.61% of storm episodes led to fatalities, while only 0.74% of events led to fatalities.
3.57% of storm episodes led to injuries, while only 1.02% of events led to injuries.
15.10% of storm episodes caused damage, while 15.39% of events causes damage.

```{r Who is most affected by weather}

fatalities %>%
  ggplot(mapping = aes(x = fatality_age)) +
    geom_histogram(binwidth = 4)

```

Going to add more details about what caused fatalities in different age groups and more - Andrew