---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
```

```{r load-data}
details = read_csv('Data/details_combined.csv')
fatalities = read_csv('Data/fatalities_combined.csv')
locations = read_csv('Data/locations_combined.csv')
```

```{r Details Cleaning}
details <- details %>%
  # Combine date elements
  unite("BEGIN_DATE", BEGIN_YEARMONTH, BEGIN_DAY, sep = "") %>% 
  mutate(
    # Convert date to a character for union
    BEGIN_DATE = as.character(BEGIN_DATE),
    # Add leading zero if the time is 3 digits
    BEGIN_TIME = case_when(
      BEGIN_TIME < 1000 ~ sprintf("%04d", BEGIN_TIME),
      TRUE ~ as.character(BEGIN_TIME)
    )
  ) %>%
  # Combine date and time into one string
  unite("BEGIN_DT", BEGIN_DATE, BEGIN_TIME, sep = " ") %>%
  mutate(
    # Convert to a datetime
    BEGIN_DT = ymd_hm(BEGIN_DT)
  )

# Remove unnecessary columns
details <- details %>%
  select(-c(EVENT_NARRATIVE, EPISODE_NARRATIVE, YEAR, MONTH_NAME))

# Rename incorrect column names
details <- details %>%
  rename(REGION = STATE,
         REGION_FIPS = STATE_FIPS)
```

```{r Fatalities (orange)}

```

