---
title: "AAAAA"
subtitle: "Presentation subtitle (if any)"
author: "Just Commit <br> Olivia Harris, Adam Laycock, Helen Miller, <br> Andrew Morris, Max Pohlman, Sarah Stewart"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---
```{r load-packages, include = FALSE}
library(knitr)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

# Dataset Background

## The Dataset
- The dataset covers 1.6 million weather events within the United States and its territories between 1996-2023.
- Collected by the National Oceanic & Atmospheric Administration (NOAA).


## What is a 'Storm Event'?
- Storm events are 'weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce.'
- Multiple storm events are grouped in storm episodes.

---
class: center, middle

# How have storm events changed over time and what is the effect on people and property?

---
# Dataset Overview (Liv)

---
# Severity of Storms (Sarah)

---
# Geospatial Correlations (Adam)

---
# (Max)

---
# (Helen)

---
# (Andrew)

---
# Summary

---
# Questions?
