---
title: "Helen Report Draft"
author: "helenrmiller"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(tidymodels)
library(janitor)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```

## Event Proportions and ENSO Status

To simplify the investigation of the impact of El Niño Southern Oscillation (ENSO) on our data, irrelevant variables were removed from the `details_clean` data set (Chunk 1). Then, using data from NOAA and NASA, a column displaying the ENSO status of each year was added and the resulting data frame was saved as `details_enso` (Chunk 2). Finally, a table of expected and observed events grouped by ENSO status was produced for initial analysis (Chunk 3). `exp_prop` acts as a null hypothesis-- assuming ENSO status has no impact on event frequency.The resulting table shows that the observed frequency is relatively similar to expected, meaning we fail to reject the null hypothesis.

```{r chunk_1}
#remove irrelevant variables for ENSO analysis
details_smaller <- details %>% 
  select(-region_fips,
         -cz_type, 
         -cz_fips, 
         -cz_timezone, 
         -wfo, 
         -source, 
         -tor_other_cz_state, 
         -tor_other_cz_fips, 
        )
```

```{r chunk_2}
#add a column for ENSO status
details_enso <- details_smaller %>% 
  mutate(enso_status = case_when(
         year(begin_dt) %in% c(1998, 
                               2003, 
                               2007, 
                               2010, 
                               2016, # 1998-2022 -> (NOAA, 2022)
                               2023 #2023 > (NASA, 2024)
                               ) ~ 'el_nino',
         year(begin_dt) %in% c(1999, 
                               2000, 
                               2008, 
                               2011, 
                               2012, 
                               2021, 
                               2022
                               ) ~ 'la_nina',
         TRUE ~ 'neutral'
  ))
```

```{r chunk_3 / table_1}
#create frequency table of ENSO events
details_enso %>% 
  group_by(enso_status) %>% 
  summarise(obs_count = n(), #observed event count
            obs_prop = n() / nrow(details_enso), #observed event proportion
            exp_prop = length(unique(year(begin_dt))) #expected event proportion
            / length(unique(year(details_enso$begin_dt)))
  )
```

## Event Types and ENSO Status
Though event frequency does not appear significantly impacted by ENSO, we were also interested in if the proportion of event types varies by ENSO status. This was explored by calculating event type proportions across all years of a common ENSO status (Chunk 4). Event types with the largest difference in proportion between El Niño and La Niña years were visualised (Chunk 5).

```{r chunk_4}
#calculate proportion in El Niño years
el_nino_prop <- details_enso %>% 
  filter(enso_status == 'el_nino') %>%
  count(event_type) %>% 
  mutate(proportion_en = n / sum(n))

#calculate proportion in La Niña years
la_nina_prop <- details_enso %>% 
  filter(enso_status == 'la_nina') %>%
  count(event_type) %>% 
  mutate(proportion_ln = n / sum(n))

#calculate proportion in neutral years
neutral_prop <- details_enso %>% 
  filter(enso_status == 'neutral') %>%
  count(event_type) %>% 
  mutate(proportion_n = n / sum(n))

#join proportion tibbles
proportions_enso <- full_join(el_nino_prop, neutral_prop, by = 'event_type') %>% 
  full_join(la_nina_prop, by = 'event_type') %>% 
  top_n((abs(proportion_en - proportion_ln)), n = 10) %>% 
  pivot_longer(cols = c(proportion_en, proportion_ln, proportion_n), 
               names_to = 'enso_status', 
               values_to = 'proportion'
               )
```

```{r chunk_5}
#reorder ENSO status for plotting
proportions_enso <- proportions_enso %>%
  mutate(enso_status = factor(enso_status, levels = c("proportion_en", "proportion_n", "proportion_ln")))

#create visualisation
proportions_enso %>% 
  ggplot(aes(fill = enso_status, 
             x = proportion, 
             y = reorder(event_type, proportion)) #make plot more intuitive
         ) +
  
  geom_col(position = 'dodge') + #separate columns -> easier to compare
  
  labs(
    x = "Proportion of Total Events in Common ENSO Status Years",
    y = "Event Type",
    title = "Event Proportion by Event Type and ENSO Status",
    subtitle = "Event types with largest difference between El Niño and La Niña years",
    fill = 'ENSO Status'
  ) +
  
  scale_fill_manual(values = c('proportion_en' = 'midnightblue', 
                               'proportion_n' = 'purple2', 
                               'proportion_ln' = 'mediumorchid2'),
    labels = c('El Niño', 'Neutral', 'La Niña')
  )
```

## Change in ENSO Statuses Over Time

As a part of initial analysis, we noted a change in the slopes of event count grouped by ENSO status (Chunk 6). Events in La Niña years appear to be  increasing in monthly frequency at a slightly greater rate than those of a different ENSO status (Chunk 7).


```{r chunk_6}
slope_exploration <- details_enso %>%
  #group by month to increase number of observations, avoid daily outliers
  group_by(floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))
  
#create visualisation (including individual observations)
slope_exploration %>% 
  ggplot(mapping = aes(x = floor_date, y = log(n), color = enso_status)) +
  
  geom_point() +
  
  geom_smooth(method = lm, se = FALSE) + #add linear model
  
  scale_color_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
    labels = c('El Niño', 'Neutral', 'La Niña')
  ) +
  
  labs(
    x = "Date",
    y = "Number of Events per Month (log-scale)",
    title = "Event Count Over Time, Grouped by ENSO Status",
    subtitle = "Includes Monthly Observations and Linear Model",
    color = "ENSO Status"
  )
```

```{r chunk_7}
#plot only linear models, excluding individual observations
slope_exploration %>% 
  ggplot(mapping = aes(x=floor_date, y = log(n), color = enso_status)) +
  
  geom_smooth(method = lm, se= FALSE) +
  
  scale_color_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
    labels = c('El Niño', 'Neutral', 'La Niña')) +
  
  labs(
    x = "Date",
    y = "Number of Events per Month (log scale)",
    title = "Event Count Over Time, Grouped by ENSO Status",
    subtitle = "Includes Only Linear Model",
    color = "ENSO Status"
  )
  
```

To test if this difference in slope is statistically significant, each linear model was bootstrapped (Chunk 8). The results of the bootstrapping procedure are displayed in the output of Chunk 9. No significant difference is apparent due to the overlap in confidence intervals, though this could be due to sample size.

```{r chunk_8}
#calculate number of events per month for each ENSO status
#store in seperate tibbles
by_month_en <- details_enso %>% 
  filter(enso_status == 'el_nino') %>% 
  group_by(en_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))

by_month_ln <- details_enso %>%
  filter(enso_status == 'la_nina') %>% 
  group_by(ln_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))

by_month_n <- details_enso %>%
  filter(enso_status == 'neutral') %>% 
  group_by(n_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))

#create bootstrapping function
bootstrapping <- function(df, floor_name){
  # set a seed
  set.seed(30305)
  floor_name <- sym(floor_name)
  
  # take 1000 bootstrap samples
  enso_boot <- bootstraps(df, times = 1000)
  
  # for each sample
  # fit a model and save output in model column
  # tidy model output and save in coef_info column 
  enso_slope_models <- enso_boot %>%
    mutate(
      model = map(splits, ~ lm(n ~ !!floor_name, data = .)),
      coef_info = map(model, tidy)
    )
    
  # un-nest coef_info (for intercept and slope)
  enso_coefs <- enso_slope_models %>%
    unnest(coef_info)
    
  # calculate 95% confidence interval
  int_pctl(enso_slope_models, coef_info)
}

#display bootstrapped slopes in a tibble
el_nino_slope <- bootstrapping(by_month_en, 'en_floor_date')
View(el_nino_slope)
la_nina_slope <- bootstrapping(by_month_ln, 'ln_floor_date')
View(la_nina_slope)
neutral_slope <- bootstrapping(by_month_n, 'n_floor_date')
View(neutral_slope)

```

```{r chunk_9}
#assign ENSO status to bootstrapped results before binding
el_nino_slope$enso_status <- "el_nino"
la_nina_slope$enso_status <- "la_nina"
neutral_slope$enso_status <- "neutral"

combined_slopes <- dplyr::bind_rows(el_nino_slope, neutral_slope, la_nina_slope)

#remove intercept error bars from bound tibble
combined_slopes %>% 
  filter(term != '(Intercept)') %>% 
  
  #create visualisation 
  ggplot(mapping = aes(x = enso_status, y = .estimate, fill = enso_status)) +
  
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) + 
  
  #assign CI values to error bar
  geom_errorbar(aes(ymin = .lower, ymax = .upper), 
                position = position_dodge(width = 0.8),
                width = 0.25
                ) +
  
  scale_fill_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
  labels = c('El Niño', 'La Niña', 'Neutral')
  ) +
  
  #correct axis labelling
  scale_x_discrete(
    labels = c('el_nino' = 'El Niño', 
               'la_nina' = 'La Niña', 
               'neutral' = 'Neutral'
               )
    ) +
  
  labs(
    x = "ENSO Status",
    y = "True Slope (of log-scaled model)",
    title = "Bootstrapped Linear Model Slopes by ENSO Status",
  ) +
  
  theme(legend.position = 'none')
```

# Citations 

Ocean Surface Topography from Space. (2024). *El Niño 2023 \| El Niño/La Niña Watch & PDO*. [online] Available at: <https://sealevel.jpl.nasa.gov/data/el-nino-la-nina-watch-and-pdo/el-nino-2023/.>

FORMAT THIS: <https://psl.noaa.gov/enso/past_events.html>
