---
title: "Max Report Draft"
author: "MaxwellPohlman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(tidymodels)
library(janitor)
library(broom)
library(patchwork)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```

# Event Type Averages and Data Limitations 

We split event types into two categories: more or less damaging/deadly (compared to the global average). This assessed whether increased fatalities and damages over time were reflected equally in both severe and less severe storms. 

The below chunk shows the steps of this categorisation process: 
```{r chunk_1}
# Rename redundant event type names
  details <- details %>%
    mutate(event_type = replace(event_type, event_type == "Hurricane (Typhoon)", "Hurricane"))

# Select relevant columns and replace NA values
details_selected <- details %>%
  select(event_type, deaths_direct, deaths_indirect, damage_total, begin_dt) %>%  
  mutate(
    year = as.numeric(format(as.Date(begin_dt, format = "%Y-%m-%d"), "%Y"))  
  ) %>%
  na.omit()

# Calculate total fatalities and damage per event, and group by event type
details_grouped <- details_selected %>%
  mutate(
    fatalities_total = deaths_direct + deaths_indirect
  ) %>%
  group_by(event_type) %>%
  summarise(
    avg_fatalities_by_type = mean(fatalities_total, na.rm = TRUE),  
    avg_damage_by_type = mean(damage_total, na.rm = TRUE)  
  ) %>%
  ungroup()

# Calculate global averages and assign groups
global_averages <- details_grouped %>%
  summarise(
    global_avg_fatalities = mean(avg_fatalities_by_type, na.rm = TRUE),  
    global_avg_damage = mean(avg_damage_by_type, na.rm = TRUE)  
  )

# Add grouping for more/less deadly and damaging based on global averages
details_grouped <- details_grouped %>%
  mutate(
    fatality_group = if_else(avg_fatalities_by_type > global_averages$global_avg_fatalities,
                             "More Deadly", "Less Deadly"),
    damage_group = if_else(avg_damage_by_type > global_averages$global_avg_damage,
                           "More Damaging", "Less Damaging")
  )
```
We used average impact instead of net impact to determine whether total fatalities and damages increased due to stronger storms or simply greater storm frequency. 

We first calculated averages for the entire dataset (1996–2023):
```{r chunk_2}
# Adjusted Graph for Fatalities (All Data)
fatality_graph <- details_grouped %>%
  ggplot(aes(x = fatality_group, 
             y = avg_fatalities_by_type, 
             fill = fatality_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("More Deadly" = "red", "Less Deadly" = "green")) +
  labs(
    title = "Average Fatalities by Group",
    subtitle = "Based on Event Type Averages Compared to Global Average",
    x = "Weather Group",  
    y = "Average Fatalities per Event Type",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.subtitle = element_text(size = 6.5),  
    axis.text.x = element_text(size = 5)  
  )

# Adjusted Graph for Damages (All Data)
damage_graph <- details_grouped %>%
  ggplot(aes(x = damage_group, 
             y = avg_damage_by_type, 
             fill = damage_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("More Damaging" = "purple", "Less Damaging" = "orange")) +
  labs(
    title = "Average Damages by Group",
    subtitle = "Based on Event Type Averages Compared to Global Average",
    x = "Weather Group", 
    y = "Average Damages per Event Type (USD)",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.subtitle = element_text(size = 6.5), 
    axis.text.x = element_text(size = 5)  
  )

# Combine  plots
combined_original <- fatality_graph + damage_graph

# Display plots
combined_original
```

This was repeated for 1996–7 and 2022–3, using two-year periods to minimize outliers. Coding these graphs involved filtering for year, recalculating averages, and regrouping event types. A sample of this is shown below: 
```{r chunk_3 / table_1}
# Filter for 1996
details_1996_7 <- details_selected %>% filter(year %in% c(1996, 1997))
details_grouped_1996_7 <- details_1996_7 %>%
  mutate(fatalities_total = deaths_direct + deaths_indirect) %>%
  group_by(event_type) %>%
  summarise(
    avg_fatalities_by_type = mean(fatalities_total, na.rm = TRUE),
    avg_damage_by_type = mean(damage_total, na.rm = TRUE)
  ) %>%
  ungroup()

# Recalculate global averages for 1996 data
global_averages_1996_7 <- details_grouped_1996_7 %>%
  summarise(
    global_avg_fatalities = mean(avg_fatalities_by_type, na.rm = TRUE),
    global_avg_damage = mean(avg_damage_by_type, na.rm = TRUE)
  )

# Add grouping for 1996 data
details_grouped_1996_7 <- details_grouped_1996_7 %>%
  mutate(
    fatality_group = if_else(avg_fatalities_by_type > global_averages_1996_7$global_avg_fatalities,
                             "More Deadly", "Less Deadly"),
    damage_group = if_else(avg_damage_by_type > global_averages_1996_7$global_avg_damage,
                           "More Damaging", "Less Damaging")
  )
```

### Labeled graph for the 1996-7 period:
```{r chunk_4, echo=FALSE}
# Adjusted Graph for 1996 Fatalities
fatality_graph_1996_7 <- details_grouped_1996_7 %>%
  ggplot(aes(x = fatality_group, 
             y = avg_fatalities_by_type, 
             fill = fatality_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  labs(title = "1996-7: Fatalities by Group", subtitle = "Data from 1996-7", 
       x = "Weather Group",  
       y = "Average Fatalities") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 5) 
  )

# Adjusted Graph for 1996 Damages
damage_graph_1996_7 <- details_grouped_1996_7 %>%
  ggplot(aes(x = damage_group, 
             y = avg_damage_by_type, 
             fill = damage_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  labs(title = "1996-7: Damages by Group", subtitle = "Data from 1996-7", 
       x = "Weather Group",  
       y = "Average Damages (USD)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 5)  
  )

# Combine plots
combined_1996_7 <- fatality_graph_1996_7 + damage_graph_1996_7

# Display plots
combined_1996_7
```

### Labeled graph for the 2022-3 period:
```{r chunk_5, echo=FALSE}
# Filter for 2022-3
details_2022_3 <- details_selected %>% filter(year %in% c(2022, 2023))
details_grouped_2022_3 <- details_2022_3 %>%
  mutate(fatalities_total = deaths_direct + deaths_indirect) %>%
  group_by(event_type) %>%
  summarise(
    avg_fatalities_by_type = mean(fatalities_total, na.rm = TRUE),
    avg_damage_by_type = mean(damage_total, na.rm = TRUE)
  ) %>%
  ungroup()

# Recalculate global averages for 2022-3 data
global_averages_2022_3 <- details_grouped_2022_3 %>%
  summarise(
    global_avg_fatalities = mean(avg_fatalities_by_type, na.rm = TRUE),
    global_avg_damage = mean(avg_damage_by_type, na.rm = TRUE)
  )

# Add grouping for 2022-3 data
details_grouped_2022_3 <- details_grouped_2022_3 %>%
  mutate(
    fatality_group = if_else(avg_fatalities_by_type > global_averages_2022_3$global_avg_fatalities,
                             "More Deadly", "Less Deadly"),
    damage_group = if_else(avg_damage_by_type > global_averages_2022_3$global_avg_damage,
                           "More Damaging", "Less Damaging")
  )

# Adjusted Graph for 2022-3 Fatalities
fatality_graph_2022_3 <- details_grouped_2022_3 %>%
  ggplot(aes(x = fatality_group, 
             y = avg_fatalities_by_type, 
             fill = fatality_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  labs(title = "2022-3: Fatalities by Group", subtitle = "Data from 2022-3", 
       x = "Weather Group",  
       y = "Average Fatalities") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 5)  
  )

# Adjusted Graph for 2022-3 Damages
damage_graph_2022_3 <- details_grouped_2022_3 %>%
  ggplot(aes(x = damage_group, 
             y = avg_damage_by_type, 
             fill = damage_group)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge", alpha = 0.7) +
    geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), 
            vjust = -0.5, size = 3) +
  labs(title = "2022-3: Damages by Group", subtitle = "Data from 2022-3", 
       x = "Weather Group",  
       y = "Average Damages (USD)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 5)  
  )

# Combine plots
combined_2022_3 <- fatality_graph_2022_3 + damage_graph_2022_3

# Display plots
combined_2022_3
```
Below are tibbles showing event types, classifications, and impact averages for the entire dataset, 1996–7, and 2022–3. This frames changes and verifies classifications. 
``` {r chunk_6, echo=FALSE}
# Exact values for the original chart
details_grouped %>%
  select(event_type, fatality_group, damage_group, avg_fatalities_by_type, avg_damage_by_type) %>%
  arrange(fatality_group, damage_group)

# Exact values for 1996-7
details_grouped_1996_7 %>%
  select(event_type, fatality_group, damage_group, avg_fatalities_by_type, avg_damage_by_type) %>%
  arrange(fatality_group, damage_group) 

# Exact values for 2022-3
details_grouped_2022_3 %>%
  select(event_type, fatality_group, damage_group, avg_fatalities_by_type, avg_damage_by_type) %>%
  arrange(fatality_group, damage_group)
```
## Changes
All averaged values decreased from 1996–7 to 2022–3. 
First, by creating and joining the year groups, we produced a tibble showing the percent and absolute change for each event type. Next, we calculated percent and absolute changes for the global averages of each period. Finally, we generated a simple percent change for each more/less category. 
``` {r chunk_7}
# Create year groups
details_grouped_1996_7 <- details_grouped_1996_7 %>%
  mutate(year_group = "1996-1997")

details_grouped_2022_3 <- details_grouped_2022_3 %>%
  mutate(year_group = "2022-2023")

# Combine the data-sets
combined_data <- full_join(
  details_grouped_1996_7, 
  details_grouped_2022_3, 
  by = "event_type", 
  suffix = c("_1996_7", "_2022_3")
)

# Calculate differences and percentage changes
combined_data <- combined_data %>%
  mutate(
    change_fatalities = avg_fatalities_by_type_2022_3 - avg_fatalities_by_type_1996_7,
    change_damage = avg_damage_by_type_2022_3 - avg_damage_by_type_1996_7,
    percent_change_fatalities = (change_fatalities / avg_fatalities_by_type_1996_7) * 100,
    percent_change_damage = (change_damage / avg_damage_by_type_1996_7) * 100
  )

# View
combined_data %>%
  select(event_type, change_fatalities, percent_change_fatalities, change_damage, percent_change_damage)

# Compare the global averages between the two groups
comparison <- tibble(
  metric = c("Fatalities", "Damages"),
  avg_1996_7 = c(global_averages_1996_7$global_avg_fatalities, global_averages_1996_7$global_avg_damage),
  avg_2022_3 = c(global_averages_2022_3$global_avg_fatalities, global_averages_2022_3$global_avg_damage),
  absolute_change = avg_2022_3 - avg_1996_7,
  percent_change = (absolute_change / avg_1996_7) * 100
)

# View
comparison

# Final category changes calculations
Percent_change_less_fatalities <- ((.02-.03)/.03)*100
Percent_change_more_fatalities <- ((.55-1.07)/1.07)*100
Percent_change_less_damages <- ((46507.11-384199.67)/384199.67)*100
Percent_change_more_damages <- ((6543117.28-7033105.21)/7033105.21)*100 

# Laying it all out
Percent_change_less_fatalities
Percent_change_more_fatalities
Percent_change_less_damages
Percent_change_more_damages
```

# Findings
While total damages and fatalities rose, average impact per event declined. This suggests that increased storm frequency has affected people and property more than storm strength.  
However, a limitation of this dataset is the inability to fully assess storm severity without accounting for infrastructure strength, which was unavailable. 


