---
title: "Liv Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
library(dplyr)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```


```{r evaluation}

# Linear regression using year to predict the number of events per year

# Making variable year
details$year <- as.numeric(format(details$begin_dt, "%Y"))

events_per_year <- aggregate(event_id ~ year, data = details, FUN = length)
names(events_per_year) <- c("year", "event_count") 

model3 <- lm(event_count ~ year, data = events_per_year)

summary(model3)

# Visualising

ggplot(events_per_year, aes(x = year, y = event_count)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", size = 1) + 
  labs(title = "Event Count Over Time",
       x = "Year",
       y = "Number of Events") +
  theme_minimal()

```



```{r Damages}

details$total_damages <- details$damage_property + details$damage_crops
model_damages <- lm(total_damages ~ year, data = details)

summary(model_damages)

damages_per_year <- details %>%
  group_by(year) %>%
  summarise(total_damages = sum(damage_property + damage_crops, na.rm = TRUE))

model_damages2 <- lm(total_damages ~ year, data = damages_per_year)

summary(model_damages2)


# Visualising

ggplot(damages_per_year, aes(x = year, y = total_damages)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", size = 1) + 
  labs(title = "Damages Total Over Time",
       x = "Year",
       y = "Total Damages") +
  theme_minimal()
```




```{r Fatalities}

# Linear Regression Total Fatalities / Year

details$fatalities_total <- details$deaths_direct + details$deaths_indirect
model_fatalities <- lm(fatalities_total ~ year, data = details)

summary(model_fatalities)

# Linear Regression Total Fatalities per Year / Year
fatalities_per_year <- details %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(deaths_direct + deaths_indirect, na.rm = TRUE))

model_fatalities2 <- lm(total_fatalities ~ year, data = fatalities_per_year)

summary(model_fatalities2)

ggplot(fatalities_per_year, aes(x = year, y = total_fatalities)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", size = 1) + 
  labs(title = "Total Fatalities Over Time",
       x = "Year",
       y = "Total Fatalities") +
  theme_minimal()

```



```{r Magnitudes}

# Group data and create model list
events_per_year <- details %>%
  group_by(event_type, year) %>%
  summarise(event_count = n())

event_types <- unique(events_per_year$event_type)

# Create an empty list to store the models
model_list <- list()

# Fit the models for each event type
for (event in event_types) {
  model_list[[event]] <- lm(event_count ~ year, data = subset(events_per_year, event_type == event))
}

# Extract the year coefficients and R-squared values
coefficients <- sapply(model_list, function(m) coef(m)["year"])
r_squared <- sapply(model_list, function(m) summary(m)$r.squared)

# Filter for events with R-squared > 0.5
valid_events <- names(coefficients[r_squared > 0.5])

# Subset coefficients for valid events
valid_coefficients <- coefficients[valid_events]

# Sort the valid coefficients in descending order
top10_coefficients <- sort(valid_coefficients, decreasing = TRUE)

# Print the top 10 events with highest coefficients
print("Top 10 increasing trends:")
print(top10_coefficients[1:10])

```


```{r event type breakdown}

events_per_year <- details %>%
  group_by(event_type, year) %>%
  summarise(event_count = n())

event_types <- unique(events_per_year$event_type)

model_list <- list()
for (event in event_types) {
  model_list[[event]] <- lm(event_count ~ year, data = subset(events_per_year, event_type == event))
}

model_list

summary(model_list[["Thunderstorm Wind"]])$r.squared

summary(model_list[["Winter Weather"]])$r.squared

summary(model_list[["Marine Thunderstorm Wind"]])$r.squared

summary(model_list[["Heavy Rain"]])$r.squared

summary(model_list[["Debris Flow"]])$r.squared

summary(model_list[["Marine High Wind"]])$r.squared

summary(model_list[["Rip Current"]])$r.squared

summary(model_list[["Lightning"]])$r.squared


# Extract the year coefficient for each event type
coefficients <- sapply(model_list, function(m) coef(m)["year"])

# Sort coefficients in descending order (top trends)
top10_coefficients <- sort(coefficients, decreasing = TRUE)
top10 <- head(top10_coefficients, 10)

# Sort coefficients in ascending order (bottom trends)
bottom10_coefficients <- sort(coefficients, decreasing = FALSE)
bottom10 <- head(bottom10_coefficients, 10)

# Print top 10 and bottom 10 coefficients directly
print("Top 10 increasing trends:")
print(top10)

print("Bottom 10 decreasing trends:")
print(bottom10)

# Finding R squared
r_squared <- sapply(model_list, function(m) summary(m)$r.squared)
valid_events <- names(coefficients[r_squared > 0.5])
print(valid_events)

```



