---
title: "Helen Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_csv('../Data/details_combined.csv')
fatalities = read_csv('../Data/fatalities_combined.csv')
```

```{r Details Cleaning}
# Merge and convert formats of begin-date variables
details <- details %>%
  # Combine date elements
  unite("BEGIN_DATE", BEGIN_YEARMONTH, BEGIN_DAY, sep = "") %>% 
  mutate(
    # Convert date to a character for union with time
    BEGIN_DATE = as.character(BEGIN_DATE),
    # Add leading zero if the time is 3 digits
    BEGIN_TIME = case_when(
      BEGIN_TIME < 1000 ~ sprintf("%04d", BEGIN_TIME),
      TRUE ~ as.character(BEGIN_TIME)
    )
  ) %>%
  # Unite date and time into one string
  unite("BEGIN_DT", BEGIN_DATE, BEGIN_TIME, sep = " ") %>%
  mutate(
    # Convert to a datetime
    BEGIN_DT = ymd_hm(BEGIN_DT)
  )

# Merge and convert formats of end-date variables
details <- details %>%
  # Combine date elements
  unite("END_DATE", END_YEARMONTH, END_DAY, sep = "") %>% 
  mutate(
    # Convert date to a character for union with time
    END_DATE = as.character(END_DATE),
    # Add leading zero if the time is 3 digits
    END_TIME = case_when(
      END_TIME < 1000 ~ sprintf("%04d", END_TIME),
      TRUE ~ as.character(END_TIME)
    )
  ) %>%
  # Unite date and time into one string
  unite("END_DT", END_DATE, END_TIME, sep = " ") %>%
  mutate(
    # Convert to a datetime
    END_DT = ymd_hm(END_DT)
  )

# Remove unnecessary columns
details <- details %>%
  select(-c(YEAR, MONTH_NAME, BEGIN_DATE_TIME, END_DATE_TIME))

# Rename incorrect column names
details <- details %>%
  rename(REGION = STATE,
         REGION_FIPS = STATE_FIPS)

# Create duration variable 
details <- details %>% 
  mutate(
    duration = END_DT - BEGIN_DT
  ) %>% 
    mutate(
      duration = as.numeric(duration) / 60
    )

# Change cost suffixes
details <- details %>% 
  mutate(
    DAMAGE_PROPERTY = as.numeric(gsub("K", "e+03", gsub("M", "e+06", DAMAGE_PROPERTY))),
    DAMAGE_CROPS = as.numeric(gsub("K", "e+03", gsub("M", "e+06", DAMAGE_CROPS)))
  )

details <- details %>%
  mutate(
    DAMAGE_TOTAL = DAMAGE_PROPERTY + DAMAGE_CROPS
  )

details <- details %>% 
  filter(
    !year(BEGIN_DT) == '2024'
  )

details <- details %>% 
  clean_names()
```

```{r Fatalities Cleaning}
fatalities <- fatalities %>% 
  select(-FAT_TIME, -FATALITY_DATE, -EVENT_YEARMONTH) %>% 
  mutate(YMD = ymd(paste(FAT_YEARMONTH, FAT_DAY))) %>% 
  select(-FAT_DAY, -FAT_YEARMONTH)

fatalities <- fatalities %>%
  filter(!year(YMD) == '2024')

fatalities <- fatalities %>% 
  clean_names()
```

```{r}
#remove irrelevant variables for ENSO analysis
details_smaller <- details %>% 
  select(-region_fips, #I can simplify this, see details cleaning above
         -cz_type, 
         -cz_fips, 
         cz_timezone, 
         -wfo, 
         -source, 
         -tor_other_cz_state, 
         -tor_other_cz_fips, 
         -episode_narrative,
         -event_narrative)

#add a column for ENSO status
details_enso <- details_smaller %>% 
  mutate(enso_status = case_when(
         year(begin_dt) %in% c(1998, 
                               2003, 
                               2007, 
                               2010, 
                               2016
                               ) ~ 'el_nino',
         year(begin_dt) %in% c(1999, 
                               2000, 
                               2008, 
                               2011, 
                               2012, 
                               2021, 
                               2022
                               ) ~ 'la_nina',
         TRUE ~ 'neutral'
  ))

details_enso %>% 
  group_by(enso_status) %>% 
  summarise(count = n())
```

Source = <https://psl.noaa.gov/enso/past_events.html>