---
title: "Helen Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```

```{r}
#remove irrelevant variables for ENSO analysis
details_smaller <- details %>% 
  select(-region_fips,
         -cz_type, 
         -cz_fips, 
         cz_timezone, 
         -wfo, 
         -source, 
         -tor_other_cz_state, 
         -tor_other_cz_fips, 
        )

#add a column for ENSO status
details_enso <- details_smaller %>% 
  mutate(enso_status = case_when(
         year(begin_dt) %in% c(1998, 
                               2003, 
                               2007, 
                               2010, 
                               2016,
                               2023
                               ) ~ 'el_nino',
         year(begin_dt) %in% c(1999, 
                               2000, 
                               2008, 
                               2011, 
                               2012, 
                               2021, 
                               2022
                               ) ~ 'la_nina',
         TRUE ~ 'neutral'
  ))

#create frequency table of ENSO events
details_enso %>% 
  group_by(enso_status) %>% 
  summarise(count = n(),
            prop = n()/nrow(details_enso),
            exp_prop = length(unique(year(begin_dt)))
            /length(unique(year(details_enso$begin_dt)))
  )
```

Source = <https://psl.noaa.gov/enso/past_events.html> added 2023 as an el nino year

