---
title: "Sarah Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```



```{r filter events}
severe_details <- details %>%
  select(-c(episode_id, wfo, source, region_fips, end_dt, region_fips, cz_type, cz_fips, cz_name, cz_timezone, injuries_direct, injuries_indirect, deaths_direct, deaths_indirect, flood_cause, category, tor_f_scale, tor_length, tor_width, tor_other_wfo, tor_other_cz_state, tor_other_cz_fips, tor_other_cz_name, begin_azimuth, end_azimuth, begin_lon, begin_lat, end_lon, end_lat, duration, magnitude_type, begin_range, begin_location, end_range)) %>%
  group_by(event_type) %>%
  mutate(sum_damage_total = sum(damage_total, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(
    desc(sum_damage_total),
  ) %>%
  distinct(event_type, .keep_all = TRUE) 
severe_details

severe_details_recent

```


```{r joining}
combined_data <- severe_details %>%
  full_join(fatalities, by = "event_id")

```



```{r damage over time}
details %>%
  mutate(
 year = year(as.Date(begin_dt))
  ) %>%
  group_by(year)%>%
    ggplot(mapping = aes(x = year, y = log(damage_total), color = factor(event_type))) + 
  geom_col() + 
  labs(x = "year", y = "total damage($)", title = "Total damage over years")

```

```{r no fatalities per region}
fatalities_by_region <-combined_data%>%
  group_by(region) %>%
  summarize(
    no_fatalities = n_distinct(fatality_id)) %>%
  mutate(no_fatalities = no_fatalities) %>%
  ungroup()

total_fatalities <- combined_data %>%
  summarize(total_fatalities = n_distinct(fatality_id)) %>%
  pull(total_fatalities)
  
fatalities_by_region <- fatalities_by_region %>%
  mutate(total_fatalities= total_fatalities)

combined_data <- combined_data %>%
  left_join(fatalities_by_region, by="region")

```



```{r fatalities frequency map}

combined_data <- combined_data %>% 
  mutate(region = tolower(region))

fatalities_by_region <- fatalities_by_region %>%
  mutate(region = tolower(region))

fatalities_by_region
             
usa_map <- map_data("state")
             
map_combined <- usa_map %>% 
  left_join(fatalities_by_region, by = "region") 


map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = no_fatalities, color = "black"
    )
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Map of the USA") +
  theme_minimal()
```
```{r frequency events}
frequency_by_region <-combined_data%>%
  group_by(region) %>%
  summarize(
    no_events = n_distinct(event_id)) %>%
  mutate(no_events = no_events) %>%
  ungroup()

total_events <- combined_data %>%
  summarize(total_events = n_distinct(event_id)) %>%
  pull(total_events)
  
frequency_by_region <- frequency_by_region %>%
  mutate(total_events= total_events)

combined_data <- combined_data %>%
  left_join(frequency_by_region, by="region")

```

```{r event frequency map}

combined_data <- combined_data %>% 
  mutate(region = tolower(region))

frequency_by_region <- frequency_by_region %>%
  mutate(region = tolower(region))

             
usa_map <- map_data("state")
             
map_combined <- usa_map %>% 
  left_join(frequency_by_region, by = "region") 


map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = no_events, color = "black"
    )
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Map of the USA") +
  theme_minimal()
```


```{r map damage}
library(ggplot2)
library(maps)

#load map data
combined_data <- combined_data %>% 
      mutate(region = tolower(region))
             
usa_map <- map_data("state")     
             
map_combined_now <- combined_data_now %>% 
  full_join(usa_map)

# Plot the map
map_combined_now %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = log(sum_damage_total)
    )
  ) +
  geom_polygon(colour='black') +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Map of the USA") +
  theme_minimal()

#https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
```


```{r damage over time2} 
overtime_severe_details <- severe_details %>%
  slice_head(n=10)

overtime_severe_details %>%
  mutate(
    year = year(as.Date(begin_dt))
  ) %>%
  filter(
    year >2000
  ) %>%
  group_by(year) %>%
  ggplot(mapping = aes(x = year, y = log(damage_total), color = factor(event_type))) +
  geom_point(size = 1) +
  facet_wrap(~event_type, scales = "free") + 
  labs(
    x = "year", y = "total damage ($)")
```
