---
title: "Helen Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(tidymodels)
library(janitor)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```

# Organise Data for ENSO Analysis

```{r}
#remove irrelevant variables for ENSO analysis
details_smaller <- details %>% 
  select(-region_fips,
         -cz_type, 
         -cz_fips, 
         cz_timezone, 
         -wfo, 
         -source, 
         -tor_other_cz_state, 
         -tor_other_cz_fips, 
        )

#add a column for ENSO status
details_enso <- details_smaller %>% 
  mutate(enso_status = case_when(
         year(begin_dt) %in% c(1998, 
                               2003, 
                               2007, 
                               2010, 
                               2016,
                               2023
                               ) ~ 'el_nino',
         year(begin_dt) %in% c(1999, 
                               2000, 
                               2008, 
                               2011, 
                               2012, 
                               2021, 
                               2022
                               ) ~ 'la_nina',
         TRUE ~ 'neutral'
  ))
```

# Explore Change in Event Numbers Over Time (Based on ENSO Status)

```{r}
slope_exploration <- details_enso %>%
  #group events by month
  group_by(floor_date = floor_date(begin_dt, 'month')) %>% 
  #find proportions that each event accounts for in different types of years
  summarise(n = n(), enso_status = unique(enso_status))
  
#plot with points
slope_exploration %>% 
  ggplot(mapping = aes(x=floor_date, y = log(n), color = enso_status)) +
  geom_point() +
  geom_smooth(method = lm, se= FALSE) +
  scale_color_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
    labels = c('El Niño', 'Neutral', 'La Niña')
  ) +
  labs(
    x="Date",
    y="Number of Events per Month (log scale)",
    title="Event Count Over Time, Grouped by ENSO Status",
    subtitle = "Includes Monthly Observations and Linear Model",
    color = 'ENSO Status'
  )
  
#plot with only lm
slope_exploration %>% 
  ggplot(mapping = aes(x=floor_date, y = log(n), color = enso_status)) +
  geom_smooth(method = lm, se= FALSE) +
  scale_color_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
    labels = c('El Niño', 'Neutral', 'La Niña')) +
  labs(
    x="Date",
    y="Number of Events per Month (log scale)",
    title="Event Count Over Time, Grouped by ENSO Status",
    subtitle = "Includes Only Linear Model",
    color = 'ENSO Status'
  )
  
```

```{r}
by_month_en <- details_enso %>%
  filter(enso_status == 'el_nino') %>% 
  group_by(en_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))

by_month_ln <- details_enso %>%
  filter(enso_status == 'la_nina') %>% 
  group_by(ln_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))

by_month_n <- details_enso %>%
  filter(enso_status == 'neutral') %>% 
  group_by(n_floor_date = floor_date(begin_dt, 'month')) %>% 
  summarise(n = n(), enso_status = unique(enso_status))
```


```{r}
bootstrapping <- function(df, floor_name){
  # set a seed
  set.seed(30305)
  floor_name <- sym(floor_name)
  
  # take 1000 bootstrap samples
  enso_boot <- bootstraps(df, times = 1000)
  
  # for each sample
  # fit a model and save output in model column
  # tidy model output and save in coef_info column 
  enso_slope_models <- enso_boot %>%
    mutate(
      model = map(splits, ~ lm(n ~ !!floor_name, data = .)),
      coef_info = map(model, tidy)
    )
    
  # unnest coef_info (for intercept and slope)
  enso_coefs <- enso_slope_models %>%
    unnest(coef_info)
    
  # calculate 95% (default) percentile interval
  int_pctl(enso_slope_models, coef_info)
}

el_nino_slope <- bootstrapping(by_month_en, 'en_floor_date')
View(el_nino_slope)
la_nina_slope <- bootstrapping(by_month_ln, 'ln_floor_date')
View(la_nina_slope)
neutral_slope <- bootstrapping(by_month_n, 'n_floor_date')
View(neutral_slope)

```

```{r}
el_nino_slope$enso_status <- "el_nino"
la_nina_slope$enso_status <- "la_nina"
neutral_slope$enso_status <- "neutral"

combined_slopes <- dplyr::bind_rows(el_nino_slope, neutral_slope, la_nina_slope)

combined_slopes %>% 
  filter(term != '(Intercept)') %>% #removes intercept error bars
  
  ggplot(mapping = aes(x = enso_status, y = .estimate, fill = enso_status)) +
  
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) + 
  
  geom_errorbar(aes
                (ymin = .lower, ymax = .upper), 
                position = position_dodge(width = 0.8),
                width = 0.25
                ) +
  
  scale_fill_manual(values = c('el_nino' = 'midnightblue',
                                'neutral' = 'purple2', 
                                'la_nina' = 'mediumorchid2'
                                ),
  labels = c('El Niño', 'La Niña', 'Neutral')
  ) +
  
  scale_x_discrete(
    labels = c('el_nino' = 'El Niño', 
               'la_nina' = 'La Niña', 
               'neutral' = 'Neutral'
               )
    ) +
  
  labs(
    x="ENSO Status",
    y="True Slope (of log-scaled model)",
    title="Bootstrapped Linear Model Slopes by ENSO Status",
  ) +
  
  theme(legend.position = 'none')
```



```{r}
#calculate proportion of total events by event type for el nino years
el_nino_freq <- details_enso %>% 
  filter(enso_status == 'el_nino') %>%
  count(event_type) %>% 
  mutate(proportion_en = n / sum(n))

#calculate proportion of total events by event type for la nina years
la_nina_freq <- details_enso %>% 
  filter(enso_status == 'la_nina') %>%
  count(event_type) %>% 
  mutate(proportion_ln = n / sum(n))

#calculate proportion of total events by event type in neutral years
neutral_freq <- details_enso %>% 
  filter(enso_status == 'neutral') %>%
  count(event_type) %>% 
  mutate(proportion_n = n / sum(n))

#join previous frequency dataframes into one
proportions_enso <- full_join(el_nino_freq, neutral_freq, by = 'event_type') %>% 
  full_join(la_nina_freq, by='event_type') %>% 
  top_n((abs(proportion_en-proportion_ln)), n = 10) %>% 
  pivot_longer(cols = c(proportion_en, proportion_ln, proportion_n), 
               names_to = 'enso_status', 
               values_to = 'proportion'
               )
#reorder ENSO status for plotting
proportions_enso <- proportions_enso %>%
  mutate(enso_status = factor(enso_status, levels = c("proportion_en", "proportion_n", "proportion_ln")))

#plot joined dataframe
proportions_enso %>% 
  ggplot(aes(fill=enso_status, 
             x = proportion, 
             y = reorder(event_type, proportion))
         ) +
  geom_col(position = 'dodge') +
  labs(
    x="Proportion of Total Events in Common ENSO Status Years",
    y="Event Type",
    title="Proportion of Total Events by Event Type and ENSO Status",
    subtitle="Event types with largest difference between El Niño and La Niña years",
    fill = 'ENSO Status'
  ) +
  scale_fill_manual(
    values = c('proportion_en' = 'midnightblue', 'proportion_n' = 'purple2', 'proportion_ln' = 'mediumorchid2' ),
    labels = c('El Niño', 'Neutral', 'La Niña')
  )
```

```{r, eval=FALSE}
#I'm going to keep working on this, but it's quite difficult and I doubt it'll
#be ready in time for the presentation

#tears of sadness
enso_knn <- details_enso %>% 
  select(event_id, 
         region, 
         event_type, 
         injuries_direct,
         deaths_direct, 
         damage_property, 
         damage_crops,
         enso_status
         ) %>% 
    drop_na()
enso_knn <- enso_knn %>% 
  sample_n(1000)

#train/test split
set.seed(70405)
enso_split <- initial_split(enso_knn)
enso_train <- training(enso_split)
enso_test  <- testing(enso_split)

enso_recipe_1 <- recipe(enso_status ~ ., 
                      data = enso_knn
                      ) %>% 
  #step_naomit(all_predictors()) %>% #ensure no predictors are missing
  step_novel(all_nominal(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_naomit(all_predictors())#create dummy variables

#choose a model
enso_mod_1 <- nearest_neighbor(engine='kknn', mode = 'classification')

#initiate workflow, add model and recipe
enso_wflow_1 <- workflow() %>% 
  add_model(enso_mod_1) %>% 
  add_recipe(enso_recipe_1)

#fit the model on the training data
enso_fit_1 <- enso_wflow_1 %>% 
  fit(data = enso_train)

#generate predictions for the test data
#bake applies the recipe to the test data to process it the same way as train
enso_pred_1 <- predict(enso_fit_1, enso_test, type = 'prob')

results <- enso_test %>% 
  bind_cols(
    enso_pred_1
  )
```

Source = <https://psl.noaa.gov/enso/past_events.html> added 2023 as an el nino year, find source for that as well. Note the caveat with events vs episodes

```{r}
#create frequency table of ENSO events
details_enso %>% 
  group_by(enso_status) %>% 
  summarise(count = n(),
            prop = n()/nrow(details_enso),
            exp_prop = length(unique(year(begin_dt)))
            /length(unique(year(details_enso$begin_dt)))
  )
```
