---
title: "Sarah Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
```

```{r load-data}
details = read_rds('../data/clean_data/details_clean.rds')
fatalities = read_rds('../data/clean_data/fatalities_clean.rds')
```



```{r filter events}
severe_details <- details %>%
  select(-c(episode_id, wfo, source, region_fips, end_dt, region_fips, cz_type, cz_fips, cz_name, cz_timezone, injuries_direct, injuries_indirect, deaths_direct, deaths_indirect, flood_cause, category, tor_f_scale, tor_length, tor_width, tor_other_wfo, tor_other_cz_state, tor_other_cz_fips, tor_other_cz_name, begin_azimuth, end_azimuth, begin_lon, begin_lat, end_lon, end_lat, duration, magnitude_type, begin_range, begin_location, end_range)) 

severe_details_recent <- severe_details %>%
  group_by(event_type) %>%
  mutate(sum_damage_total = sum(damage_total, na.rm = TRUE),
         year = year(as.Date(begin_dt))) %>%
  ungroup() %>%
  arrange(
    desc(sum_damage_total),
    desc(year)
  ) %>%
  distinct(event_type, .keep_all = TRUE) 

severe_details_old <- severe_details %>%
  group_by(event_type) %>%
  mutate(sum_damage_total = sum(damage_total, na.rm = TRUE),
         year = year(as.Date(begin_dt))) %>%
  ungroup() %>%
  arrange(
    desc(sum_damage_total),
    year
  ) %>%
  distinct(event_type, .keep_all = TRUE) 

severe_details_recent

```


```{r joining}
combined_data_now <- severe_details_recent %>%
  full_join(fatalities, by = "event_id")

combined_data_old <- severe_details_old%>%
  full_join(fatalities, by = "event_id")
```



```{r damage over time}
details %>%
  mutate(
 year = year(as.Date(begin_dt))
  ) %>%
  group_by(year)%>%
    ggplot(mapping = aes(x = year, y = log(damage_total), color = factor(event_type))) + 
  geom_col() + 
  labs(x = "year", y = "total damage($)", title = "Total damage over years")

```


```{r map damage}
library(ggplot2)
library(maps)

#load map data
combined_data_now <- combined_data_now %>% 
      mutate(region = tolower(region))
             
usa_map <- map_data("state")     
             
map_combined_now <- combined_data_now %>% 
  full_join(usa_map)

# Plot the map
map_combined_now %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = log(sum_damage_total), color = "black"
    )
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Map of the USA") +
  theme_minimal()

#https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
```

```{r overtime}

combined_data_old <- combined_data_old %>% 
      mutate(region = tolower(region))
             
usa_map <- map_data("state")     

map_combined_old <- combined_data_old %>% 
  full_join(usa_map)


map_combined_old %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = log(damage_total), color = "black"
    )
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Map of the USA") +
  theme_minimal()
```


```{r map magnitude}
map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = magnitude),
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "lightblue", high = "darkblue")
  labs(title = "Map of the USA") +
  theme_minimal()

```

```{r damage over time2} 
overtime_severe_details <- severe_details %>%
  slice_head(n=10)

overtime_severe_details %>%
  mutate(
    year = year(as.Date(begin_dt))
  ) %>%
  filter(
    year >2000
  ) %>%
  group_by(year) %>%
  ggplot(mapping = aes(x = year, y = log(damage_total), color = factor(event_type))) +
  geom_point(size = 1) +
  facet_wrap(~event_type, scales = "free") + 
  labs(
    x = "year", y = "total damage ($)")
```