---
title: "Liv Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
library(dplyr)
```

```{r load-data}
details = read_rds('../../data/clean_data/details_clean.rds')
fatalities = read_rds('../../data/clean_data/fatalities_clean.rds')
```


```{r evaluation}

# Linear regression using year to predict the number of events per year

# Making variable year
details$year <- as.numeric(format(details$begin_dt, "%Y"))

events_per_year <- aggregate(event_id ~ year, data = details, FUN = length)
names(events_per_year) <- c("year", "event_count") 

model3 <- lm(event_count ~ year, data = events_per_year)

summary(model3)

# Visualising

ggplot(events_per_year, aes(x = year, y = event_count)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "purple2", linetype = "solid", size = 1) + 
  labs(title = "Event Count Over Time",
       x = "Year",
       y = "Number of Events") +
  theme_minimal()

```



```{r Damages}

details$total_damages <- details$damage_property + details$damage_crops
model_damages <- lm(total_damages ~ year, data = details)

summary(model_damages)

damages_per_year <- details %>%
  group_by(year) %>%
  summarise(total_damages = sum(damage_property + damage_crops, na.rm = TRUE))

model_damages2 <- lm(total_damages ~ year, data = damages_per_year)

summary(model_damages2)


# Visualising

ggplot(damages_per_year, aes(x = year, y = total_damages)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "purple2", linetype = "solid", size = 1) + 
  labs(title = "Damages Total Over Time",
       x = "Year",
       y = "Total Damages") +
  theme_minimal()
```




```{r Fatalities}

# Linear Regression Total Fatalities / Year

details$fatalities_total <- details$deaths_direct + details$deaths_indirect
model_fatalities <- lm(fatalities_total ~ year, data = details)

summary(model_fatalities)

# Linear Regression Total Fatalities per Year / Year
fatalities_per_year <- details %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(deaths_direct + deaths_indirect, na.rm = TRUE))

model_fatalities2 <- lm(total_fatalities ~ year, data = fatalities_per_year)

summary(model_fatalities2)

ggplot(fatalities_per_year, aes(x = year, y = total_fatalities)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "purple2", linetype = "solid", size = 1) + 
  labs(title = "Total Fatalities Over Time",
       x = "Year",
       y = "Total Fatalities") +
  theme_minimal()

```



```{r event type breakdown}

# Event type linear regression brreakdown

events_per_year <- details %>%
  group_by(event_type, year) %>%
  summarise(event_count = n())

event_types <- unique(events_per_year$event_type)

model_list <- list()
for (event in event_types) {
  model_list[[event]] <- lm(event_count ~ year, data = subset(events_per_year, event_type == event))
}

model_list


# Finding Specific R- squared

summary(model_list[["Thunderstorm Wind"]])$r.squared

summary(model_list[["Winter Weather"]])$r.squared

summary(model_list[["Marine Thunderstorm Wind"]])$r.squared

summary(model_list[["Heavy Rain"]])$r.squared

summary(model_list[["Debris Flow"]])$r.squared

summary(model_list[["Marine High Wind"]])$r.squared

summary(model_list[["Rip Current"]])$r.squared

summary(model_list[["Lightning"]])$r.squared


# Extract the year coefficient for each event type
coefficients <- sapply(model_list, function(m) coef(m)["year"])

# Top 10
top10_coefficients <- sort(coefficients, decreasing = TRUE)
top10 <- head(top10_coefficients, 10)

print(top10)


# Bottom 10
bottom10_coefficients <- sort(coefficients, decreasing = FALSE)
bottom10 <- head(bottom10_coefficients, 10)

print(bottom10)

# R squared's above 0.5
r_squared <- sapply(model_list, function(m) summary(m)$r.squared)
valid_events <- names(coefficients[r_squared > 0.5])

print(valid_events)

```



