---
title: "Sarah Personal Investigation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, message = FALSE}
library(tidyverse)
library(janitor)
library(maps)
```

```{r load-data}
details = read_rds('../../data/clean_data/details_clean.rds')
fatalities = read_rds('../../data/clean_data/fatalities_clean.rds')
```


```{r filter events}
severe_details <- details %>%
  select(-c(episode_id, wfo, source, region_fips, end_dt, region_fips, cz_type, cz_fips, cz_name, cz_timezone, injuries_direct, injuries_indirect, deaths_direct, deaths_indirect, flood_cause, category, tor_f_scale, tor_length, tor_width, tor_other_wfo, tor_other_cz_state, tor_other_cz_fips, tor_other_cz_name, begin_azimuth, end_azimuth, begin_lon, begin_lat, end_lon, end_lat, duration, magnitude_type, begin_range, begin_location, end_range)) %>%
  group_by(event_type) %>%
  mutate(sum_damage_total = sum(damage_total, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(
    desc(sum_damage_total),
  ) %>%
  distinct(event_type, .keep_all = TRUE) 


```

```{r not severe}
not_severe_details <- details %>%
  select(-c(episode_id, wfo, source, region_fips, end_dt, region_fips, cz_type, cz_fips, cz_name, cz_timezone, injuries_direct, injuries_indirect, deaths_direct, deaths_indirect, flood_cause, category, tor_f_scale, tor_length, tor_width, tor_other_wfo, tor_other_cz_state, tor_other_cz_fips, tor_other_cz_name, begin_azimuth, end_azimuth, begin_lon, begin_lat, end_lon, end_lat, duration, magnitude_type, begin_range, begin_location, end_range)) %>%
  group_by(event_type) %>%
  mutate(sum_damage_total = sum(damage_total, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(
    desc(sum_damage_total),
  ) 

```


```{r joining}
combined_data <- severe_details %>%
  full_join(fatalities, by = "event_id")

```

```{r not combined}
not_combined_data <- not_severe_details %>%
  full_join(fatalities, by = "event_id")
```


```{r no fatalities per region}
fatalities_by_region <-not_combined_data%>%
  group_by(region) %>%
  summarize(
    no_fatalities = n_distinct(fatality_id)) %>%
  mutate(no_fatalities = no_fatalities) %>%
  ungroup()

total_fatalities <- not_combined_data %>%
  summarize(total_fatalities = n_distinct(fatality_id)) %>%
  pull(total_fatalities)
  
fatalities_by_region <- fatalities_by_region %>%
  mutate(total_fatalities= total_fatalities)

not_combined_data <- not_combined_data %>%
  left_join(fatalities_by_region, by="region")

```


```{r not fatalities  map}

not_combined_data <- not_combined_data %>% 
  mutate(region = tolower(region))

fatalities_by_region <- fatalities_by_region %>%
  mutate(region = tolower(region))

             
usa_map <- map_data("state")
             
map_combined <- usa_map %>% 
  left_join(fatalities_by_region, by = "region") 

map_combined

map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = no_fatalities)
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red", name = "Number of fatalities") + 
  labs(
    title = "Number of fatalities per state") +
  theme_minimal()
```



```{r not severe frequency events}
not_frequency_by_region <-not_combined_data%>%
  group_by(region) %>%
  summarize(
    no_events = n_distinct(event_id)) %>%
  mutate(no_events = no_events) %>%
  ungroup()

not_total_events <- not_combined_data %>%
  summarize(total_events = n_distinct(event_id)) %>%
  pull(total_events)
  
not_frequency_by_region <- not_frequency_by_region %>%
  mutate(not_total_events= total_events)

not_combined_data <- not_combined_data %>%
  left_join(not_frequency_by_region, by="region")

```

```{r not severe event frequency map}
not_combined_data <- not_combined_data %>% 
  mutate(region = tolower(region))

not_frequency_by_region <- not_frequency_by_region %>%
  mutate(region = tolower(region))

             
usa_map <- map_data("state")
             
map_combined <- usa_map %>% 
  left_join(not_frequency_by_region, by = "region") 


map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = no_events)
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red", name = "Number of events") + 
  labs(
    title = "Number of all weather events experienced per state") +
  theme_minimal()
```

```{r  frequency events}
frequency_by_region <-combined_data%>%
  group_by(region) %>%
  summarize(
    no_events = n_distinct(event_id)) %>%
  mutate(no_events = no_events) %>%
  ungroup()

total_events <- combined_data %>%
  summarize(total_events = n_distinct(event_id)) %>%
  pull(total_events)
  
frequency_by_region <- frequency_by_region %>%
  mutate(total_events= total_events)

combined_data <- combined_data %>%
  left_join(frequency_by_region, by="region")

```


```{r event frequency map}

combined_data <- combined_data %>% 
  mutate(region = tolower(region))

frequency_by_region <- frequency_by_region %>%
  mutate(region = tolower(region))

             
usa_map <- map_data("state")
             
map_combined <- usa_map %>% 
  left_join(frequency_by_region, by = "region") 


map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = no_events)
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red", name = "Number of severe events") + 
  labs(
    title = "Map of the USA",
    subtitle = "Most severe weather events experienced per state") +
  theme_minimal()
```

```{r map damage}

#load map data
combined_data <- combined_data %>% 
      mutate(region = tolower(region))
             
usa_map <- map_data("state")     
             
map_combined<- combined_data%>% 
  full_join(usa_map)

# Plot the map
map_combined %>% 
  ggplot(
    mapping = aes(
      x = long, y = lat, group = group, fill = log(sum_damage_total)
    )
  ) +
  geom_polygon() +
  scale_fill_gradient(low = "yellow", high = "red", name = "Total Damage ($)") + 
  labs(title = "States experiencing the most damage") +
  theme_minimal()



#https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
```





